---
topic: "air"
title: "Status of Fine Particulate Matter in B.C. (2014-2016)"
output: envreportbc::print_ver
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, 
                      error = FALSE, fig.width = 7)

library("rcaaqs")
library("readr")
library("dplyr")
library("tidyr")
library("sf")
library("bcmaps")

library("ggplot2")
library("grid")

library("envreportutils")
library("extrafont")

load("../tmp/analysed.RData")
knitr::read_chunk("../04_output.R")

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}
```

```{r pre}

```


British Columbia operates a network of air monitoring stations that
measure fine particulate matter and other air pollutants. This indicator
reports on the concentration of fine particulate matter from 2011-2013
and compares it to the [Canadian Ambient Air Quality Standards
(CAAQS)](http://www.ccme.ca/en/current_priorities/air/caaqs.html)
developed by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

\begin{center}
  \fbox{
    \begin{minipage}[c]{12cm}
      \centering{\textbf Fine Particulate Matter\\ Canadian Ambient Air Quality Standards (CAAQS)}
    
      \begin{itemize}
          \itemsep1pt\parskip0pt\parsep0pt
        \item
          The \textbf{PM$_{\text{2.5}}$ 24-hour metric} should be lower than \textbf{28 µg/m3}. The
          PM$_{\text{2.5}}$ 24-hour metric is the 3-year average of the annual 98th percentile
          of the daily 24-hour average concentration.
        \item
          The \textbf{PM$_{\text{2.5}}$ annual metric} should be lower than \textbf{10.0 µg/m3}. The
          PM$_{\text{2.5}}$ annual metric is the 3-year average of the annual average
          concentration.
      \end{itemize}
      \href{http://www.ccme.ca/en/current_priorities/air/caaqs.html}{Learn 
            more about CAAQS fine particulate matter metric calculations.}
    \end{minipage}
  }
\end{center}

-   **Fine particulate matter is an air pollutant.** Solid or liquid
    particles floating in the air are called particulate matter. The
    smallest of these particles - those that are 2.5 microns or less in
    diameter, or less than 1/20th the width of a human hair - are called
    fine particulate matter. Fine particulate matter comes from many
    natural and human activities, including wildfires and emissions from
    prescribed burning, forestry operations, residential woodstoves, and
    transportation^[1](#fn1)^.
-   **Fine particulate matter can be harmful to humans.** Exposure to
    fine particulate matter has been associated with several serious
    health effects including heart and lung disease^[1](#fn1)^. Both
    short-term (24-hour) and longer-term (a year or more) exposure to
    fine particulate matter can have negative effects on human health.
-   **Fine particulate matter levels met the standard for the CAAQS
    measures at 46 of the 50 assessed monitoring stations in B.C.** The two
    PM~2.5~ CAAQS metrics exceeded the Canadian Ambient Air Quality
    Standard at two of the 50 stations (4%). The PM~2.5~ 24-hour metric
    exceeded the standard at an additional two stations (4%). Across all
    50 stations the PM~2.5~ 24-hour metric ranged from 6 to 36 micrograms
    per cubic meter of air (µg/m^3^). The PM~2.5~ annual metric ranged
    from 1.6 to 10.8 µg/m^3^.
-   **Two of B.C.'s seven air zones met the Canadian Ambient Air Quality 
    Standard for both fine particulate matter metrics.** One air zone met the 
    standard for the annual metric only and two air zones exceeded the standards 
    for both metrics. There was not enough data to estimate the CAAQS PM~2.5~ 
    metrics for the remaining two air zones for this reporting period.
-   **CAAQS fine particulate matter metrics are used to set management
    levels for each air zone.** Four management levels (green, yellow,
    orange, and red) are each associated with a suite of actions that
    become more rigorous as fine particulate matter metric values
    approach the Canadian Ambient Air Quality Standards.
    
The maps and chart below summarise the Canadian Ambient Air Quality Standard
(CAAQS) achievement status for fine particulate matter in B.C. air zones, as
well as both the annual and 24-hour CAAQS PM~2.5~ metric values at individual
monitoring stations. Summaries are given for each monitoring station where
sufficient data was available for the 2011-2013 reporting period.

```{r achievement_map_24}
```

```{r fig.height=8}
plot(achievement_map_24)
```

```{r achievement_map_annual}
```

```{r fig.height=8}
plot(achievement_map_annual)
```

```{r summary_plot}
```

```{r fig.height=10}
plot(ambient_summary_plot)
```


#### More about the CAAQS fine particulate matter metrics: 

-   There are two CAAQS metrics for fine particulate matter (PM~2.5~):
    PM~2.5~ 24-hour, 3-year average of the annual 98th percentile of the
    daily 24-hour average concentrations of PM~2.5~, and PM~2.5~ annual, the
    3-year average of the annual average concentration of PM~2.5~.
-   PM~2.5~ levels for an air zone are the highest CAAQS PM~2.5~ metric
    values reported from monitoring stations within the air zone. However, 
    stations with metric values based on only two years of data are excluded 
    for consideration of the *air zone* metric value.
-   Only air monitoring stations with sufficient data on fine
    particulate matter concentrations for the 2011-2013 reporting period
    were included in this indicator. Visit [BC Air
    Quality](http://www.bcairquality.ca/readings/index.html) for a
    complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the CAAQS PM~2.5~
    metrics are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf)
    
### Canada-wide Air Quality Management System

The Air Quality Management System (AQMS) is Canada's approach for
protecting air quality. Under the AQMS, provinces and territories
monitor, report and manage local air quality within air
zones, with the goal of continuously improving air quality and
keeping pollutants below the Canadian Ambient Air Quality Standards
(CAAQS).

-   Provinces and territories establish air zones that are defined by a
    similar set of air quality characteristics, issues and trends.
-   There are seven air zones in British Columbia: Coastal, Georgia
    Strait, Lower Fraser Valley, Southern Interior, Central Interior,
    Northeast and Northwest.
-   The Air Zone Management Framework defines management levels based on
    the highest annual and 24-hour CAAQS fine particulate matter metric values 
    reported from monitoring stations within the air zone. Concluding air zone 
    management levels for PM~2.5~ are based on the higher of the PM~2.5~ annual and 
    the PM~2.5~ 24-hour management levels.
-   The four management levels are (1) green (actions for keeping clean
    areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing CAAQS
    exceedance); and (4) red (actions for achieving air zone CAAQS).
    Each management level is associated with a suite of actions that
    become more rigorous as fine particulate matter concentration levels
    approach the CAAQS.
-   Based on the highest annual and 24-hour CAAQS fine particulate matter 
    metric values reported within each air zone, the Central Interior and 
    Georgia Strait air zones were assigned to a red management level (actions 
    for achieving air zone CAAQS); the Lower Fraser Valley air zones was 
    assigned to a yellow management level (actions for preventing air quality 
    deterioration); the Southern Interior air zone to an orange management level
    (actions for preventing CAAQS exceedance); and the Coastal air zone
    was assigned to a green air zone (actions for keeping clean areas
    clean). Due to insufficient data, particulate management levels in
    the Northwest and Northeast air zones are yet to be determined.
-   Actions for each management level are detailed in the CCME [Guidance
    Document on Air Zone Management
    (2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).
    
```{r mgmt_map}
```

```{r fig.height=9}
mgmt_map + 
  labs(title = "Air Quality Management Levels in B.C. Air Zones") +
  theme(legend.position = "bottom", legend.direction = "vertical")
```


```{r mgmt_chart}
```

```{r fig.height=8}
mgmt_chart + 
  labs(title = "Station-level Air Quality Management Levels Within Air Zones")
```

#### More about the AQMS management levels:

-   There can be cases when an air zone does not achieve a given CAAQS
    because of sources over which jurisdictions have little or no
    control, such as those related to transboundary flows and
    exceptional events like forest fires.
-   Before determining management levels, jurisdictions have the option
    of adjusting their air zone metric values to remove such external
    influences. These arrangements aim to ensure that jurisdictions are
    responsible for managing only the emissions sources they can
    control.
-   In B.C., wildfires are generally the largest contributor to
    transboundary flows and exceptional events. However, the highest
    fine particulate matter levels for the 2011-2013 reporting period
    occurred outside of the historic wildfire season. Thus, the
    exceedances of the PM2.5 CAAQS are not believed to be a result of
    transboundary flows and exceptional events. Air zone management
    levels for fine particulate matter were thus determined directly
    from the CAAQS fine particulate matter metrics with no adjustments
    for this reporting period.

### Methods

The methods used to develop this indicator- including procedures, data
requirements, and calculation of the CAAQS fine particulate matter
metrics-are detailed in the [Guidance Document on Achievement
Determination: Canadian Ambient Air Quality Standards for Fine
Particulate Matter and Ozone
(2012)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf)<span
class="fileinfo"> (PDF)</span> published by the [Canadian Council of
Ministers of the Environment (CCME)](http://www.ccme.ca/index.html). **R package and code:** We have developed an [R](http://www.r-project.org) package to facilitate the calculation of air quality metrics according to the Canadian Ambient Air Quality Standards [CAAQS](http://www.ccme.ca/en/current_priorities/air/caaqs.html). Download the 'rcaaqs' package from [GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating the analysis presented on this page is also available on [GitHub](https://github.com/bcgov/pm25_caaqs_analysis).

### References and Other Useful Links

Visit [BC Air Quality](http://www.bcairquality.ca/plans/national-air-quality-management-system.html) for more information on the implementation of the AQMS in B.C., and to read individual Air Zone reports on the achievement of CAAQS for ground-level ozone and fine particulate matter.

Access [B.C.'s Air Quality
Readings](http://www.bcairquality.ca/readings/index.html) for real-time
data on air pollutants and locations of all air monitoring stations in
B.C.

For more details on the CAAQS and Canada's Air Quality Management System
visit [Canada-wide Air Quality Management System
(AQMS)](http://www.ccme.ca/en/resources/air/aqms.html)

[BC Lung Association's BC State of the Air
Reports](http://www.bc.lung.ca/airquality/stateoftheair-report.html).

[Canadian Environmental Sustainability Indicators: Air and Climate
Indicators](%20http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)

^1^[B.C. Air Action Plan (June
2008)](http://www.bcairsmart.ca/docs/bcairactionplan.pdf)<span
class="fileinfo"> (PDF, 1.1MB)</span>

### Data

\*By accessing these datasets, you agree to the license associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Fine Particulate Matter CAAQS 2011-2013](http://catalogue.data.gov.bc.ca/dataset/bc-ambient-air-quality-standards-reporting-results-fine-particulate-matter)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [BC PM2.5
    Hourly](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [BC Air Monitoring
    Stations](http://catalogue.data.gov.bc.ca/dataset/air-quality-monitoring-verified-hourly-data-and-station-data)
    (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
-   [BC Air Zones](http://catalogue.data.gov.bc.ca/dataset/british-columbia-air-zones) (License: [B.C.
    OGL](http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf))
    
Published and Available On-Line at Environmental Reporting BC (August 2015):
<http://www.env.gov.bc.ca/soe/indicators/air/fine_pm.html>

Email correspondence to: envreportbc@gov.bc.ca

# Appendix: Detailed CAAQS results for each monitoring station within air zones in B.C.

```{r stn_plots, message=FALSE}

```


```{r appendix, results='asis', fig.height=3}

pm_data <- sf::st_set_geometry(pm_stats, NULL)

for (az in unique(pm_data$Airzone)) {
  emsids <- unique(pm_data$ems_id[pm_stats$Airzone == az])
  cat("\\subsection{", az, " Air Zone}", sep = "")

  for (emsid in emsids) {
    
    caaqs_data_24h <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_24h")
    caaqs_data_annual <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_annual")
    
    site <- pm_stats$station_name[pm_data$ems_id == emsid][1]

    ## Subset daily, caaqs, and annual data
    daily_plot <- stnplots[[emsid]]$daily
    annual_plot <- stnplots[[emsid]]$annual

    cat("\\subsubsection*{\\underline{", knitr_latex_char(site), " monitoring station}}", sep = "")

    cat("\\textbf{PM2.5 24-hour Air Quality Standard:} ",
        as.character(caaqs_data_24h$caaqs), " \\newline ")
    cat("\\textbf{PM2.5 24-hour Metric:} ", caaqs_data_24h$metric_value,
        " µg/m³ (", caaqs_data_24h$n_years,
        " year average) \\newline ", sep = "")
    plot(daily_plot)

    cat("\\newline ")
    cat("\\textbf{PM2.5 Annual Air Quality Standard:} ",
        as.character(caaqs_data_annual$caaqs), " \\newline ")
    cat("\\textbf{PM2.5 Annual Metric:} ", caaqs_data_annual$metric_value,
        " µg/m³ (", caaqs_data_annual$n_years,
        " year average) \\newline ", sep = "")
    plot(annual_plot)

    cat("\\newpage ")
  }
}

```

