---
topic: "air"
title: "Status of Fine Particulate Matter in B.C. (2014-2016)"
output: envreportbc::print_ver
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, 
                      error = FALSE, fig.width = 7)

library("rcaaqs")
library("readr")
library("dplyr")
library("tidyr")
library("sf")
library("bcmaps")

library("ggplot2")
library("grid")

library("envreportutils")
library("extrafont")

load("../tmp/analysed.RData")
knitr::read_chunk("../04_output.R")

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}
```

```{r pre}

```


British Columbia operates a network of air monitoring stations that
measure fine particulate matter and other air pollutants. This indicator
reports on the concentration of fine particulate matter from 2014-2016
and compares it to the [Canadian Ambient Air Quality
Standard](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

\begin{center}
  \fbox{
    \begin{minipage}[c]{12cm}
      \centering{\textbf Fine Particulate Matter\\ Canadian Ambient Air Quality Standards for PM$_{\text{2.5}}$}
    
      \begin{itemize}
          \itemsep1pt\parskip0pt\parsep0pt
        \item
          The \textbf{PM$_{\text{2.5}}$ 24-hour metric} should be lower than \textbf{28 µg/m3}.
        \item
          The \textbf{PM$_{\text{2.5}}$ annual metric} should be lower than \textbf{10.0 µg/m3}.
      \end{itemize}
      \href{http://www.ccme.ca/en/current_priorities/air/caaqs.html}{Learn 
            more about CAAQS fine particulate matter metric calculations.}
    \end{minipage}
  }
\end{center}

-   **Fine particulate matter is an air pollutant.** Solid or liquid
    particles floating in the air are called particulate matter. The
    smallest of these particles - those that are 2.5 microns or less in
    diameter, or less than 1/20th the width of a human hair - are called
    fine particulate matter (PM2.5). Fine particulate matter comes from
    many natural and human activities, including wildfires and emissions
    from prescribed burning, forestry operations, residential
    woodstoves, and transportation.^[1](#fn1)^
-   **Fine particulate matter can be harmful to humans.** Exposure to
    fine particulate matter has been associated with several serious
    health effects including heart and lung disease.^[1](#fn1)^ Both
    short-term (24-hour) and longer-term (a year or more) exposure to
    fine particulate matter can have negative effects on human health.
-   **Fine particulate matter levels met the Canadian Ambient Air
    Quality Standard at 86 percent of assessed monitoring stations in
    B.C.** There are two standards for PM2.5 - an annual standard, and a
    24-hour standard (see sidebar). The annual standard was met at all
    of the 49 stations for which valid data was obtained, while the
    24-hour standard was met at 43 of the 50 stations (86%) with
    sufficient data for analysis.
-   **Two of B.C.'s seven air zones met the Canadian Ambient Air Quality
    Standard for both fine particulate matter metrics.** All air zones
    met the standard for the annual metric, and three air zones exceeded
    the standards for the 24-hour metric. There was not enough data to
    estimate the PM2.5 metrics Northeast air zones for this reporting
    period. Currently, there are no [air monitoring
    stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest air zone.
-   **Fine particulate matter metrics are used to set management levels
    for each air zone.** Four management levels (green, yellow, orange,
    and red) are each associated with a suite of actions that become
    more rigorous as fine particulate matter levels approach the
    Canadian Ambient Air Quality Standards.
    
The maps and chart below summarise the Canadian Ambient Air Quality Standard
(CAAQS) achievement status for fine particulate matter in B.C. air zones, as
well as both the annual and 24-hour CAAQS PM~2.5~ metric values at individual
monitoring stations. Summaries are given for each monitoring station where
sufficient data was available for the 2014-2016 reporting period.

```{r achievement_map_24}
```

```{r fig.height=8}
plot(achievement_map_24)
```

```{r achievement_map_annual}
```

```{r fig.height=8}
plot(achievement_map_annual)
```

```{r summary_plot}
```

```{r fig.height=10}
plot(ambient_summary_plot)
```


### More about the Canadian Ambient Air Quality Standard fine particulate matter metrics:

-   There are two standards for fine particulate matter (PM2.5): the
    24-hour standard, which is the 3-year average of the annual 98th
    percentile of the daily average concentrations of PM2.5, and the
    annual standard, which is the 3-year average of the annual average
    concentration of PM2.5.
-   The PM2.5 metric for an air zone is the highest PM2.5 metric value
    reported from monitoring stations within the air zone. However,
    stations with metric values based on only two years of data are
    excluded for consideration of the *air zone* metric value, unless 
    all station metric values are based on 2 years of data.
-   Only air monitoring stations with sufficient data on fine
    particulate matter concentrations for the 2014-2016 reporting period
    were included in this indicator. Visit [Current Air Quality
    Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    for a complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the two PM2.5 metrics
    are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).
    
### Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting
air quality. Under the Air Quality Management System, provinces and
territories monitor, report and manage local air quality within air
zones, with the goal of continuously improving air quality and keeping
pollutants below the Canadian Ambient Air Quality Standards.

-   Provinces and territories establish air zones that are defined by a
    similar set of air quality characteristics, issues and trends. There
    are seven air zones in British Columbia: Coastal, Georgia Strait,
    Lower Fraser Valley, Southern Interior, Central Interior, Northeast
    and Northwest.
-   The Air Zone Management Framework defines air zone management levels
    based on the highest annual and 24-hour fine particulate matter
    metric values reported from monitoring stations within the air zone.
    Air zone management levels for PM2.5 are based on the higher of the
    PM2.5 annual and the PM2.5 24-hour management levels.
-   The four management levels are (1) green (actions for keeping clean
    areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing Canadian Ambient
    Air Quality Standard exceedance); and (4) red (actions for achieving
    air zone Canadian Ambient Air Quality Standard). Each management
    level is associated with a suite of actions that become more
    rigorous as fine particulate matter concentration levels approach
    the standard. Actions for each management level are detailed in the
    CCME [Guidance Document on Air Zone Management (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf)
-   The Central Interior and Georgia Strait air zones were assigned to a 
    red management level; the Lower Fraser Valley and Coastal air zones were 
    assigned to a yellow management level; and the Southern Interior to an 
    orange management level. Due to insufficient data, particulate management 
    levels in the Northwest and Northeast air zones are yet to be determined. 
    Note that while two stations in the Coastal air zone (Kitimat Haisla 
    Village and Kitimat Smeltersite Road) had an annual metric that would 
    indicate an 'orange' management level, they were excluded from consideration 
    of the air zone metric value because they are based on only two years of 
    data. Please see the box below for information on how influences from 
    wildfire smoke is accounted for in the determination of management levels. 
    
```{r mgmt_map}
```

```{r fig.height=9}
mgmt_map + 
  labs(title = "Air Quality Management Levels in B.C. Air Zones") +
  theme(legend.position = "bottom", legend.direction = "vertical")
```


```{r mgmt_chart}
```

```{r fig.height=8}
mgmt_chart + 
  labs(title = "Station-level Air Quality Management Levels Within Air Zones")
```

#### More about the AQMS management levels:

-   There can be cases when an air zone does not achieve a given
    standard because of sources over which jurisdictions have little or
    no control, such as those related to transboundary flows and
    exceptional events like forest fires.
-   Before determining management levels, jurisdictions have the option
    of adjusting their air zone metric values to remove such external
    influences. These arrangements aim to ensure that jurisdictions are
    responsible for managing only the emissions sources they can
    control.
-   In B.C., wildfires are generally the largest contributor to
    transboundary flows and exceptional events. Over the 2014-16
    reporting period, 50 stations had one or more days during wildfire
    season (May-September) where the daily fine particulate matter
    concentration was determined to be influenced by wildfire and
    exceeded the standard of 28 Âµg/m^3^. Those daily concentrations were
    removed from the calculations of management levels. The majority of
    wildfire influences occurred in 2015, where fine particulate
    concentrations were affected by smoke at 44 monitoring stations for
    periods spanning one to ten days. In 2014 the fires were more
    localized, and 15 stations had up to 14 days affected by wildfire
    smoke, while in 2016 only two stations were affected.

### Methods

The methods used to develop this indicator - including procedures, data
requirements, and calculation of the Canadian Ambient Air Quality
Standard fine particulate matter metrics - are detailed in the [Guidance
Document on Achievement Determination: Canadian Ambient Air Quality
Standards for Fine Particulate Matter and Ozone (2012)
(PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf)
published by the [Canadian Council of Ministers of the Environment
(CCME)](http://www.ccme.ca/index.html).

**R package and code:** We have developed an
[**R**](http://www.r-project.org) package to facilitate the calculation
of air quality metrics according to the [Canadian Ambient Air Quality
Standards](http://www.ccme.ca/en/resources/air/aqms.html#anchor1).
Download the 'rcaaqs' package from
[GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating
the analysis presented on this page is also available on
[GitHub](https://github.com/bcgov/pm25-caaqs-analysis).

### References and Other Useful Links

-   [Learn more about the implementation of the Air Quality Management
    System in British
    Columbia](http://www2.gov.bc.ca/gov/content?id=8FA84ADC845A49AB93E525A58D9D4A97)
-   Read individual [Air Zone
    reports](http://www2.gov.bc.ca/gov/search?q=air+zone%2Binmeta%3AnrsAirQuality_nrs_air_quality_topic%3DAir+Zone+Reports&id=04D14E7F7EE043889E1AC20E205CA3F8&tab=1&sourceId=267EE7E2432B4DE89BF1905EF80F505E)
    on the achievement of the Canadian Ambient Air Quality Standards for
    ground-level ozone and fine particulate matter in B.C.
-   [Access B.C.'s real-time data on air pollutants and find locations
    of all provincial air monitoring stations in
    B.C.](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
-   [BC Lung Association's BC State of the Air
    Reports](https://bc.lung.ca/state-air-report)
-   [Canadian Environmental Sustainability Indicators: Air and Climate
    Indicators](%20http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)
-   For more details on the Canadian Ambient Air Quality Standars and
    Canada's Air Quality Management System visit [Canada-wide Air
    Quality Management
    System](http://www.ccme.ca/en/resources/air/aqms.html)
-   [Canadian Smog Science Assessment: Highlights and Key Messages](http://publications.gc.ca/site/eng/9.694820/publication.html) (PDF, 1.1MB)

### Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Fine Particulate Matter Canadian Ambient Air
    Quality Standards
    2014-2016](https://catalogue.data.gov.bc.ca/dataset/699be99e-a9ba-403e-b0fe-3d13f84f45ab) (`r get_data_license("ogl-bc")`)
-   [BC PM2.5 Hourly Data & Air Monitoring Station
    locations](https://catalogue.data.gov.bc.ca/dataset/77eeadf4-0c19-48bf-a47a-fa9eef01f409) (`r get_data_license("ogl-bc")`)
-   [BC Air
    Zones](https://catalogue.data.gov.bc.ca/dataset/e8eeefc4-2826-47bc-8430-85703d328516) (`r get_data_license("ogl-bc")`)

Published and Available On-Line at Environmental Reporting BC (August 2017):
<http://www.env.gov.bc.ca/soe/indicators/air/fine_pm.html>

Email correspondence to: envreportbc@gov.bc.ca

# Appendix: Detailed CAAQS results for each monitoring station within air zones in B.C.

```{r stn_plots, message=FALSE}

```


```{r appendix, results='asis', fig.height=3}

## Function to identify the layers in the plot that are labels ("annotations")
label_layers <- function(plot) {
  layer_is_label <- vapply(plot$layers, function(x) {
    "label" %in% names(x$aes_params)}
    , FUN.VALUE = logical(1))
  which(layer_is_label)
}

## Function to set the text size of labels in the plot
set_label_size <- function(plot, size) {
  layers <- label_layers(plot)
  
  for (l in layers) {
    plot$layers[[l]]$aes_params$size <- size
  }
  plot
}

pm_data <- sf::st_set_geometry(pm_stats, NULL)

for (az in unique(pm_data$Airzone)) {
  emsids <- unique(pm_data$ems_id[pm_stats$Airzone == az])
  cat("\\subsection{", az, " Air Zone}", sep = "")
  
  for (emsid in emsids) {
    
    caaqs_data_24h <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_24h")
    caaqs_data_annual <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_annual")
    
    site <- pm_stats$station_name[pm_data$ems_id == emsid][1]
    
    ## Subset daily, caaqs, and annual data
    daily_plot <- stnplots[[emsid]]$daily
    annual_plot <- stnplots[[emsid]]$annual
    
    cat("\\subsubsection*{\\underline{", knitr_latex_char(site), " monitoring station}}", sep = "")
    
    if (nrow(caaqs_data_24h)) {
      
      cat("\\textbf{PM2.5 24-hour Air Quality Standard:} ",
          as.character(caaqs_data_24h$caaqs), " \\newline ")
      cat("\\textbf{PM2.5 24-hour Metric:} ", caaqs_data_24h$metric_value,
          " µg/m³ (", caaqs_data_24h$n_years,
          " year average) \\newline ", sep = "")
      plot(set_label_size(daily_plot, 3))
      
    } else {
      cat("\\textbf{PM2.5 24-hour Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newline ")
    cat("\\newline ")
    cat("\\newline ")
    
    if (nrow(caaqs_data_annual)) {
      
      cat("\\textbf{PM2.5 Annual Air Quality Standard:} ",
          as.character(caaqs_data_annual$caaqs), " \\newline ")
      cat("\\textbf{PM2.5 Annual Metric:} ", caaqs_data_annual$metric_value,
          " µg/m³ (", caaqs_data_annual$n_years,
          " year average) \\newline ", sep = "")
      plot(set_label_size(annual_plot, 3))
      
    } else {
      cat("\\textbf{PM2.5 Annual Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newpage ")
  }
}

```

