---
topic: "air"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{siunitx}
  - \sisetup{detect-all,per-mode=symbol}
output: envreportutils.internal::print_ver
params:
  min_year: 2016
  max_year: 2018
title: "Status of Fine Particulate Matter in B.C. (2016-2018)"
---
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, message = FALSE, 
                      warning = FALSE, error = FALSE, fig.width = 7)

library("rcaaqs")
library("readr")
library("dplyr")
library("tidyr")
library("sf")
library("bcmaps")

library("ggplot2")
library("grid")

library("envreportutils")
library("extrafont")

load("../tmp/analysed.RData")
knitr::read_chunk("../04_output.R")

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}

year_range <- paste(params$min_year, params$max_year, sep = "-")
```

```{r pre}
# run the @knitr pre chunk in ../04_output.R
```


British Columbia operates a network of air monitoring stations that
measure fine particulate matter and other air pollutants. This indicator
reports on the concentration of fine particulate matter from `r year_range`
and compares it to the [Canadian Ambient Air Quality
Standards](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

\begin{center}

\fbox{
  \begin{minipage}[c]{15cm}
    
      \begingroup\centering{\textbf{Fine Particulate Matter (PM$_{\text{2.5}}$): Canadian Ambient Air Quality Standard}}
      
      \endgroup
      
      \vspace{1mm}
          
      PM$_{\text{2.5}}$ levels are calculated using \textbf{two} \hyperref[metric-info]{statistical forms called the annual and the 24-hour metrics}. The Canadian Ambient Air Quality Standard value for PM$_{\text{2.5}}$ annual is \textbf{10.0 micrograms per cubic metre} and the standard value for PM$_{\text{2.5}}$ 24-hour is \textbf{28 micrograms per cubic metre (\SI{}{\micro\gram\per\metre\cubed})}.
          
      \vspace{2mm}
      
      PM$_{\text{2.5}}$ standards are achieved when annual levels are $\leq$\SI{10}{\micro\gram\per\metre\cubed} and 24-hour levels are $\leq$\SI{28}{\micro\gram\per\metre\cubed}.
          
  \end{minipage}
}


\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{What is an Air Zone?}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{British Columbia is divided up into seven air zones, which are areas that typically exhibit similar air}
    
    \text{quality characteristics, issues, and trends.}
    
  \end{minipage}
}
\end{center}
\vspace{2mm}

-   **Fine particulate matter is an air pollutant.** Solid or liquid
    particles floating in the air are called particulate matter. The
    smallest of these particles---those that are 2.5 microns or less in
    diameter, or less than 1/20th the width of a human hair---are called
    fine particulate matter (PM~2.5~). Fine particulate matter comes from
    many natural and human activities, including wildfires and emissions
    from prescribed burning, forestry operations, residential
    woodstoves, and transportation.^[1](#fn1)^
-   **Fine particulate matter can be harmful to humans.** Exposure to
    fine particulate matter has been associated with several serious
    health effects including heart and lung disease.^[1](#fn1)^ Both
    short-term (24-hour) and longer-term (a year or more) exposure to
    fine particulate matter can have negative effects on human health.
-   **Fine particulate matter levels met the Canadian Ambient Air
    Quality Standards at `r min(station_summary_annual$percent_achieved, station_summary_24h$percent_achieved)`
    percent of assessed monitoring stations in B.C.** There are two standards 
    for PM~2.5~---an annual standard and a
    24-hour standard (see sidebar). The annual standard was met at 
    `r station_summary_annual$n_achieved` of the `r station_summary_annual$n`
    stations (
    `r station_summary_annual$percent_achieved`%) for 
    which valid data was obtained, while the 24-hour standard was met at 
    `r station_summary_24h$n_achieved` of the `r station_summary_24h$n` stations (
    `r station_summary_24h$percent_achieved`%) with sufficient data for analysis.
-   **Fine particulate matter levels met both of the Canadian Ambient
    Air Quality Standards in two of B.C.'s seven air zones.** The Coastal, 
    Northeast, Lower Fraser Valley, and Georgia Strait air zones met the
    PM~2.5~ annual standard, while the Central Interior and
    Southern Interior air zones exceeded it. The Coastal and Northeast 
    air zones met the PM~2.5~ 24-hour standard, while the 
    Central Interior, Southern Interior, Lower Fraser Valley, and Georgia
    Strait air zones exceeded it. Many of the exceedances were due to the
    influence of smoke from wildfires, especially during the summers of 2017 and 2018. 
    Currently, there are no 
    [air monitoring stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest air zone.
-   **Fine particulate matter levels are used to set management levels
    for each air zone.** Four management levels (green, yellow, orange,
    and red) are each associated with a suite of actions that become
    more rigorous as fine particulate matter levels approach the
    Canadian Ambient Air Quality Standards.

The maps and chart below summarise the Canadian Ambient Air Quality Standard 
achievement status for fine particulate matter in B.C. air zones, as
well as both the annual and 24-hour PM~2.5~ levels at individual
monitoring stations. Summaries are given for each monitoring station where
sufficient data was available for the `r year_range` reporting period.

# Status of 24-hour PM~2.5~ Levels in B.C. Air Zones (`r year_range`)

```{r achievement_map_24h}
# run the @knitr achievement_map_24h chunk in ../04_output.R
```

```{r, fig.height=8}
achievement_map_24h
```


\newpage

# Status of Annual PM~2.5~ Levels in B.C. Air Zones (`r year_range`)

```{r achievement_map_annual}
# run the @knitr achievement_map_annual chunk in ../04_output.R
```

```{r, fig.height=8}
achievement_map_annual
```

\newpage

# PM~2.5~ Metrics for Air Monitoring Statons within B.C. Air Zones (`r year_range`)

```{r pm_ambient_summary_plot}
# run the @knitr pm_ambient_summary_plot chunk in ../04_output.R
```
```{r, fig.height=9}
pm_ambient_summary_plot
```

\hyperdef{}{metric-info}{}
### More about the Canadian Ambient Air Quality Standard fine particulate matter metrics:

-   There are two Canadian Ambient Air Quality Standards for fine
    particulate matter (PM~2.5~). The statistical form of the PM~2.5~ annual
    metric is the 3-year average of the annual average concentration of
    PM~2.5~. The statistical form of the PM~2.5~ 24-hour metric is the
    3-year average of the annual 98th percentile of the daily average
    concentrations of PM~2.5~.
-   The PM~2.5~ metric for an air zone is the highest PM~2.5~ metric value
    reported from monitoring stations within the air zone. Air zone
    metric values are only taken from stations reporting on two years of
    data when there are no available three-year reporting stations in
    the air zone.
-   Only air monitoring stations with sufficient data on fine
    particulate matter concentrations for the `r year_range` reporting period
    were included in this indicator. Visit [Current Air Quality
    Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    for a complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the two PM~2.5~ metrics
    are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).
    
### Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting
air quality. Under the Air Quality Management System, provinces and
territories monitor, report and manage local air quality within air
zones, with the goal of continuously improving air quality and keeping
pollutants below the Canadian Ambient Air Quality Standards.

-   There are seven air zones in British Columbia: Coastal, Georgia
    Strait, Lower Fraser Valley, Southern Interior, Central Interior,
    Northeast and Northwest.
-   The Air Zone Management Framework defines air zone management levels
    for fine particulate matter based on the higher of the annual and
    24-hour fine particulate matter levels reported from monitoring
    stations within the air zone.
-   The four management levels are (1) green (actions for keeping clean
    areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing Canadian Ambient
    Air Quality Standard exceedance); and (4) red (actions for achieving
    air zone Canadian Ambient Air Quality Standard). Each management
    level is associated with a suite of actions that become more
    rigorous as fine particulate matter levels approach the standard.
    Actions for each management level are detailed in the [Guidance
    Document on Air Zone Management (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).
-   The Central Interior and Georgia Strait air zones were assigned to a
    red management level, the Southern Interior and Lower Fraser Valley to an orange management
    level, and the Coastal, and Northeast air zones
    were assigned to a yellow management level. There are currently no 
    [air monitoring stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest air zone.
    
\newpage
    
# Air Quality Management Levels in B.C. Air Zones   

```{r pm_mgmt_map}
# run the @knitr pm_mgmt_map chunk in ../04_output.R
```

```{r, fig.height=9}
pm_mgmt_map + 
  theme(legend.position = "bottom", legend.direction = "vertical")
```

\newpage

# Station-Level Air Quality Management Levels Within Air Zones

```{r pm_mgmt_chart}
# run the @knitr pm_mgmt_chart chunk in ../04_output.R
```

```{r, fig.height=8}
pm_mgmt_chart
```

#### More about the Air Quality Management System management levels:

-   There can be cases when an air zone does not achieve a given
    Canadian Ambient Air Quality Standard because of sources over which
    jurisdictions have little or no control, such as those related to
    transboundary flows and exceptional events, like forest fires.
-   Prior to determining management levels, jurisdictions have the
    option of adjusting their air zone metric values to remove such
    external influences. These arrangements aim to ensure that
    jurisdictions are responsible for managing only the emissions
    sources they can control.
-   Wildfires are generally the largest contributor to transboundary
    flows and exceptional events in B.C. Over the `r year_range` reporting
    period, **59** stations had one or more days during the wildfire season
    (May-September) where the daily fine particulate matter
    concentration was determined to be influenced by smoke from
    wildfires and exceeded the 24-hour standard of \SI{28}{\micro\gram\per\metre\cubed}. The
    majority of wildfire influences occurred in 2017. Those daily
    concentrations were removed from the calculations of management
    levels, resulting in a change from red to orange for the Southern
    Interior management level, and a change from red to yellow for the 
    Lower Fraser Valley management level for fine particulate matter.

### Methods

The methods used to develop this indicator---including procedures, data
requirements, and calculation of the Canadian Ambient Air Quality
Standard fine particulate matter metrics---are detailed in the [Guidance
Document on Achievement Determination: Canadian Ambient Air Quality
Standards for Fine Particulate Matter and Ozone (2012)
(PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf)
published by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

**R package and code:** We have developed an
[**R**](http://www.r-project.org) package to facilitate the calculation
of air quality metrics according to the [Canadian Ambient Air Quality
Standards](http://www.ccme.ca/en/resources/air/aqms.html#anchor1).
Download the 'rcaaqs' package from
[GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating
the analysis presented on this page is also available on
[GitHub](https://github.com/bcgov/pm25-caaqs-analysis).

### References and Other Useful Links

-   Read individual [Air Zone
    reports](http://www2.gov.bc.ca/gov/search?id=04D14E7F7EE043889E1AC20E205CA3F8&tab=1&sourceId=267EE7E2432B4DE89BF1905EF80F505E&q=%22air+zone+report%22)
    on the achievement of the Canadian Ambient Air Quality Standards for
    ground-level ozone and fine particulate matter in B.C.
-   [Learn more about the implementation of the Air Quality Management
    System in British
    Columbia](http://www2.gov.bc.ca/gov/content?id=8FA84ADC845A49AB93E525A58D9D4A97)
-   [Access B.C.'s real-time data on air pollutants and find locations
    of all provincial air monitoring stations in
    B.C.](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
-   [BC Lung Association's BC State of the Air
    Reports](https://bc.lung.ca/state-air-report)
-   [Canadian Environmental Sustainability Indicators: Air and Climate
    Indicators](%20http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)
-   <div id="fn1">

    </div>

    ^1^[Canadian Smog Science Assessment: Highlights and Key
    Messages](http://publications.gc.ca/site/eng/9.694820/publication.html)

### Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Fine Particulate Matter Canadian Ambient Air
    Quality Standards
    `r year_range`](https://catalogue.data.gov.bc.ca/dataset/699be99e-a9ba-403e-b0fe-3d13f84f45ab) (`r get_data_license("ogl-bc")`)
-   [BC PM~2.5~ Hourly Data & Air Monitoring Station
    locations](https://catalogue.data.gov.bc.ca/dataset/77eeadf4-0c19-48bf-a47a-fa9eef01f409) (`r get_data_license("ogl-bc")`)
-   [BC Air
    Zones](https://catalogue.data.gov.bc.ca/dataset/e8eeefc4-2826-47bc-8430-85703d328516) (`r get_data_license("ogl-bc")`)

Published and Available On-Line at Environmental Reporting BC (December 2019):
<http://www.env.gov.bc.ca/soe/indicators/air/fine_pm.html>

Email correspondence to: envreportbc@gov.bc.ca

*Suggested Citation*:  
Environmental Reporting BC. 2019. Status of Fine Particulate Matter in B.C. (`r year_range`). State of Environment Reporting, Ministry of Environment and Climate Change Strategy, British Columbia, Canada.

# Appendix: Detailed fine particulate matter Candadian Ambient Air Quality Standard results for each monitoring station within air zones in B.C.

```{r stn_plots}
# run the @knitr stn_plots chunk in ../04_output.R
```


```{r message=FALSE, results='asis', fig.height=3}

## Function to identify the layers in the plot that are labels ("annotations")
label_layers <- function(plot) {
  layer_is_label <- vapply(plot$layers, function(x) {
    "label" %in% names(x$aes_params)}
    , FUN.VALUE = logical(1))
  which(layer_is_label)
}

## Function to set the text size of labels in the plot
set_label_size <- function(plot, size) {
  layers <- label_layers(plot)
  
  for (l in layers) {
    plot$layers[[l]]$aes_params$size <- size
  }
  plot
}

pm_data <- pm_caaqs_stations_all

for (az in unique(pm_data$airzone)) {
  emsids <- unique(pm_data$ems_id[pm_data$airzone == az])
  cat("\\subsection{", az, " Air Zone}", sep = "")
  
  for (emsid in emsids) {
    
    caaqs_data_24h <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_24h")
    caaqs_data_annual <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_annual")
    
    site <- pm_data$station_name[pm_data$ems_id == emsid][1]
    
    ## Subset daily, caaqs, and annual data
    daily_plot <- stn_plots[[emsid]]$daily
    annual_plot <- stn_plots[[emsid]]$annual
    
    cat("\\subsubsection*{\\underline{", knitr_latex_char(site), " monitoring station}}", sep = "")
    
    if (nrow(caaqs_data_24h)) {
      
      cat("\\textbf{PM$_{2.5}$ 24-hour Air Quality Standard:} ",
          as.character(caaqs_data_24h$caaqs_ambient), " \\newline ")
      cat("\\textbf{PM$_{2.5}$ 24-hour Metric:} \\SI{", caaqs_data_24h$metric_value_ambient, 
          "}{\\micro\\gram\\per\\metre\\cubed} (", caaqs_data_24h$n_years,
          " year average) \\newline ", "\\newline ", sep = "")
      plot(set_label_size(daily_plot, 3))
      
    } else {
      cat("\\textbf{PM$_{2.5}$ 24-hour Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newline ")
    cat("\\newline ")
    cat("\\newline ")
    
    if (nrow(caaqs_data_annual)) {
      
      cat("\\textbf{PM$_{2.5}$ Annual Air Quality Standard:} ",
          as.character(caaqs_data_annual$caaqs_ambient), " \\newline ")
      cat("\\textbf{PM$_{2.5}$ Annual Metric:} \\SI{", caaqs_data_annual$metric_value_ambient,
          "}{\\micro\\gram\\per\\metre\\cubed} (", caaqs_data_annual$n_years,
          " year average) \\newline ",  "\\newline ", sep = "")
      plot(set_label_size(annual_plot, 3))
      
    } else {
      cat("\\textbf{PM$_{2.5}$ Annual Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newpage ")
  }
}

```

