---
topic: "air"
title: "Status of Fine Particulate Matter in B.C. (2014-2016)"
output: envreportbc::print_ver
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, 
                      error = FALSE, fig.width = 7)

library("rcaaqs")
library("readr")
library("dplyr")
library("tidyr")
library("sf")
library("bcmaps")

library("ggplot2")
library("grid")

library("envreportutils")
library("extrafont")

load("../tmp/analysed.RData")
knitr::read_chunk("../04_output.R")

knitr_latex_char <- function(x) {
  y <- gsub("\\\\", "\\\\textbackslash{}", x) # backslash has to be first!
  y <- gsub("([#$%&_])", "\\\\\\1", y) # Doesdn't deal with { or } because of function{}
  y <- gsub("\\^", "\\\\textasciicircum{}", y)
  y <- gsub("~", "\\\\textasciitilde{}", y)
  return(y)
}
```

```{r pre}

```


British Columbia operates a network of air monitoring stations that
measure fine particulate matter and other air pollutants. This indicator
reports on the concentration of fine particulate matter from 2014-2016
and compares it to the [Canadian Ambient Air Quality
Standards](http://www.ccme.ca/en/resources/air/aqms.html#anchor1)
established by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

\begin{center}

\fbox{
  \begin{minipage}[c]{15cm}
    
      \begingroup\centering{\textbf{Fine Particulate Matter (PM$_{\text{2.5}}$): Canadian Ambient Air Quality Standard}}
      
      \endgroup
      
      \vspace{1mm}
          
      PM$_{\text{2.5}}$ levels are calculated using \textbf{two} \hyperref[metric-info]{statistical forms called the annual and the 24-hour metrics}. The Canadian Ambient Air Quality Standard value for PM$_{\text{2.5}}$ annual is \textbf{10.0 micrograms per cubic metre} and the standard value for PM$_{\text{2.5}}$ 24-hour is \textbf{28 micrograms per cubic metre (µg/m\textsuperscript{3})}.
          
      \vspace{2mm}
      
      PM$_{\text{2.5}}$ standards are achieved when annual levels are $\leq$10 µg/m\textsuperscript{3} and 24-hour levels are $\leq$28 µg/m\textsuperscript{3}.
          
  \end{minipage}
}


\fbox{
  \begin{minipage}[c]{15cm}
  
    \begingroup\centering{\textbf{What is an Air Zone?}}
    
    \endgroup
    
    \vspace{1mm}
    
    \text{British Columbia is divided up into seven air zones, which are areas that typically exhibit similar air}
    
    \text{quality characteristics, issues, and trends.}
    
  \end{minipage}
}
\end{center}

-   **Fine particulate matter is an air pollutant.** Solid or liquid
    particles floating in the air are called particulate matter. The
    smallest of these particles—those that are 2.5 microns or less in
    diameter, or less than 1/20th the width of a human hair—are called
    fine particulate matter (PM$_{\text{2.5}}$). Fine particulate matter comes from
    many natural and human activities, including wildfires and emissions
    from prescribed burning, forestry operations, residential
    woodstoves, and transportation.^[1](#fn1)^
-   **Fine particulate matter can be harmful to humans.** Exposure to
    fine particulate matter has been associated with several serious
    health effects including heart and lung disease.^[1](#fn1)^ Both
    short-term (24-hour) and longer-term (a year or more) exposure to
    fine particulate matter can have negative effects on human health.
-   **Fine particulate matter levels met the Canadian Ambient Air
    Quality Standards at 86 percent of assessed monitoring stations in
    B.C.** There are two standards for PM$_{\text{2.5}}$—an annual standard and a
    24-hour standard (see sidebar). The annual standard was met at all
    of the 48 stations for which valid data was obtained, while the
    24-hour standard was met at 42 of the 49 stations (86%) with
    sufficient data for analysis.
-   **Fine particulate matter levels met both of the Canadian Ambient
    Air Quality Standards in two of B.C.'s seven air zones.** All air
    zones met the PM$_{\text{2.5}}$ annual standard, and three air zones exceeded
    the PM$_{\text{2.5}}$ 24-hour standard. One of the air zone PM$_{\text{2.5}}$ 24-hour
    exceedances—the Southern Interior air zone— was due to the influence
    of smoke from wildfires. There was not enough data to estimate PM$_{\text{2.5}}$
    levels for the Northeast air zone for this reporting period.
    Currently, there are no [air monitoring
    stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest air zone.
-   **Fine particulate matter levels are used to set management levels
    for each air zone.** Four management levels (green, yellow, orange,
    and red) are each associated with a suite of actions that become
    more rigorous as fine particulate matter levels approach the
    Canadian Ambient Air Quality Standards.

The maps and chart below summarise the Canadian Ambient Air Quality Standard 
achievement status for fine particulate matter in B.C. air zones, as
well as both the annual and 24-hour PM$_{\text{2.5}}$ levels at individual
monitoring stations. Summaries are given for each monitoring station where
sufficient data was available for the 2014-2016 reporting period.

# Status of 24-hour PM$_{\text{2.5}}$ Levels in B.C. Air Zones (2014-2016)

```{r achievement_map_24}
```

```{r fig.height=8}
plot(achievement_map_24)
```

\newpage

# Status of Annual PM$_{\text{2.5}}$ Levels in B.C. Air Zones (2014-2016)

```{r achievement_map_annual}
```

```{r fig.height=8}
plot(achievement_map_annual)
```

\newpage

# PM$_{\text{2.5}}$ Metrics for Air Monitoring Statons within B.C. Air Zones (2014-2016)

```{r summary_plot}
```

```{r fig.height=9}
plot(ambient_summary_plot)
```

\hyperdef{}{metric-info}{}
### More about the Canadian Ambient Air Quality Standard fine particulate matter metrics:

-   There are two Canadian Ambient Air Quality Standards for fine
    particulate matter (PM$_{\text{2.5}}$). The statistical form of the PM$_{\text{2.5}}$ annual
    metric is the 3-year average of the annual average concentration of
    PM$_{\text{2.5}}$. The statistical form of the PM$_{\text{2.5}}$ 24-hour metric is the
    3-year average of the annual 98th percentile of the daily average
    concentrations of PM$_{\text{2.5}}$.
-   The PM$_{\text{2.5}}$ metric for an air zone is the highest PM$_{\text{2.5}}$ metric value
    reported from monitoring stations within the air zone. Air zone
    metric values are only taken from stations reporting on two years of
    data when there are no available three-year reporting stations in
    the air zone.
-   Only air monitoring stations with sufficient data on fine
    particulate matter concentrations for the 2014-2016 reporting period
    were included in this indicator. Visit [Current Air Quality
    Data](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    for a complete list of air monitoring stations across B.C.
-   Data completeness and sufficiency criteria for the two PM$_{\text{2.5}}$ metrics
    are described in the [Guidance Document on Achievement
    Determination: Canadian Ambient Air Quality Standards for Fine
    Particulate Matter and Ozone (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf).
    
### Canada-wide Air Quality Management System

The Air Quality Management System is Canada's approach for protecting
air quality. Under the Air Quality Management System, provinces and
territories monitor, report and manage local air quality within air
zones, with the goal of continuously improving air quality and keeping
pollutants below the Canadian Ambient Air Quality Standards.

-   There are seven air zones in British Columbia: Coastal, Georgia
    Strait, Lower Fraser Valley, Southern Interior, Central Interior,
    Northeast and Northwest.
-   The Air Zone Management Framework defines air zone management levels
    for fine particulate matter based on the higher of the annual and
    24-hour fine particulate matter levels reported from monitoring
    stations within the air zone.
-   The four management levels are (1) green (actions for keeping clean
    areas clean); (2) yellow (actions for preventing air quality
    deterioration); (3) orange (actions for preventing Canadian Ambient
    Air Quality Standard exceedance); and (4) red (actions for achieving
    air zone Canadian Ambient Air Quality Standard). Each management
    level is associated with a suite of actions that become more
    rigorous as fine particulate matter levels approach the standard.
    Actions for each management level are detailed in the [Guidance
    Document on Air Zone Management (2012)
    (PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1481_gdazm_e.pdf).
-   The Central Interior and Georgia Strait air zones were assigned to a
    red management level, the Lower Fraser Valley and Coastal air zones
    were assigned to a yellow management level, and the Southern
    Interior to an orange management level. There was not enough data to
    estimate the management level for the Northeast air zone for this
    reporting period, and currently there are no [air monitoring
    stations](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
    in the Northwest air zone.
    
\newpage
    
# Air Quality Management Levels in B.C. Air Zones   
    
```{r mgmt_map}
```

```{r fig.height=9}
mgmt_map + 
  theme(legend.position = "bottom", legend.direction = "vertical")
```

\newpage

# Station-Level Air Quality Management Levels Within Air Zones

```{r mgmt_chart}
```

```{r fig.height=8}
plot(mgmt_chart) 
```

#### More about the Air Quality Management System management levels:

-   There can be cases when an air zone does not achieve a given
    Canadian Ambient Air Quality Standard because of sources over which
    jurisdictions have little or no control, such as those related to
    transboundary flows and exceptional events, like forest fires.
-   Prior to determining management levels, jurisdictions have the
    option of adjusting their air zone metric values to remove such
    external influences. These arrangements aim to ensure that
    jurisdictions are responsible for managing only the emissions
    sources they can control.
-   Wildfires are generally the largest contributor to transboundary
    flows and exceptional events in B.C. Over the 2014-16 reporting
    period, 50 stations had one or more days during the wildfire season
    (May-September) where the daily fine particulate matter
    concentration was determined to be influenced by smoke from
    wildfires and exceeded the 24-hour standard of 28 µg/m^3^. The
    majority of wildfire influences occurred in 2015. Those daily
    concentrations were removed from the calculations of management
    levels, resulting in a change from red to orange for the Southern
    Interior management level.
-   Note that while the Kitimat Haisla Village station in the Coastal
    air zone had an annual level that would indicate an 'orange'
    management level, it was excluded from consideration because the
    level was based on only two years of data, with three-year reporting
    stations available in the air zone.

### Methods

The methods used to develop this indicator—including procedures, data
requirements, and calculation of the Canadian Ambient Air Quality
Standard fine particulate matter metrics—are detailed in the [Guidance
Document on Achievement Determination: Canadian Ambient Air Quality
Standards for Fine Particulate Matter and Ozone (2012)
(PDF)](http://www.ccme.ca/files/Resources/air/aqms/pn_1483_gdad_eng.pdf)
published by the [Canadian Council of Ministers of the
Environment](http://www.ccme.ca/index.html).

**R package and code:** We have developed an
[**R**](http://www.r-project.org) package to facilitate the calculation
of air quality metrics according to the [Canadian Ambient Air Quality
Standards](http://www.ccme.ca/en/resources/air/aqms.html#anchor1).
Download the 'rcaaqs' package from
[GitHub](https://github.com/bcgov/rcaaqs). The source code for repeating
the analysis presented on this page is also available on
[GitHub](https://github.com/bcgov/pm25-caaqs-analysis).

### References and Other Useful Links

-   Read individual [Air Zone
    reports](http://www2.gov.bc.ca/gov/search?id=04D14E7F7EE043889E1AC20E205CA3F8&tab=1&sourceId=267EE7E2432B4DE89BF1905EF80F505E&q=%22air+zone+report%22)
    on the achievement of the Canadian Ambient Air Quality Standards for
    ground-level ozone and fine particulate matter in B.C.
-   [Learn more about the implementation of the Air Quality Management
    System in British
    Columbia](http://www2.gov.bc.ca/gov/content?id=8FA84ADC845A49AB93E525A58D9D4A97)
-   [Access B.C.'s real-time data on air pollutants and find locations
    of all provincial air monitoring stations in
    B.C.](http://www2.gov.bc.ca/gov/content?id=292EC5BD62A149A192B7DD2CE155E33E)
-   [BC Lung Association's BC State of the Air
    Reports](https://bc.lung.ca/state-air-report)
-   [Canadian Environmental Sustainability Indicators: Air and Climate
    Indicators](%20http://www.ec.gc.ca/indicateurs-indicators/default.asp?lang=En&n=03603FB3-1)
-   <div id="fn1">

    </div>

    ^1^[Canadian Smog Science Assessment: Highlights and Key
    Messages](http://publications.gc.ca/site/eng/9.694820/publication.html)

### Data

\*By accessing these datasets, you agree to the licence associated with
each file, as indicated in parentheses below.

-   [Indicator data: BC Fine Particulate Matter Canadian Ambient Air
    Quality Standards
    2014-2016](https://catalogue.data.gov.bc.ca/dataset/699be99e-a9ba-403e-b0fe-3d13f84f45ab) (`r get_data_license("ogl-bc")`)
-   [BC PM$_{\text{2.5}}$ Hourly Data & Air Monitoring Station
    locations](https://catalogue.data.gov.bc.ca/dataset/77eeadf4-0c19-48bf-a47a-fa9eef01f409) (`r get_data_license("ogl-bc")`)
-   [BC Air
    Zones](https://catalogue.data.gov.bc.ca/dataset/e8eeefc4-2826-47bc-8430-85703d328516) (`r get_data_license("ogl-bc")`)

Published and Available On-Line at Environmental Reporting BC (September 2017):
<http://www.env.gov.bc.ca/soe/indicators/air/fine_pm.html>

Email correspondence to: envreportbc@gov.bc.ca

# Appendix: Detailed fine particulate matter Candadian Ambient Air Quality Standard results for each monitoring station within air zones in B.C.

```{r stn_plots, message=FALSE}

```


```{r appendix, results='asis', fig.height=3}

## Function to identify the layers in the plot that are labels ("annotations")
label_layers <- function(plot) {
  layer_is_label <- vapply(plot$layers, function(x) {
    "label" %in% names(x$aes_params)}
    , FUN.VALUE = logical(1))
  which(layer_is_label)
}

## Function to set the text size of labels in the plot
set_label_size <- function(plot, size) {
  layers <- label_layers(plot)
  
  for (l in layers) {
    plot$layers[[l]]$aes_params$size <- size
  }
  plot
}

pm_data <- sf::st_set_geometry(pm_stats, NULL)

for (az in unique(pm_data$Airzone)) {
  emsids <- unique(pm_data$ems_id[pm_stats$Airzone == az])
  cat("\\subsection{", az, " Air Zone}", sep = "")
  
  for (emsid in emsids) {
    
    caaqs_data_24h <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_24h")
    caaqs_data_annual <- filter(pm_data, ems_id == !!emsid, metric == "pm2.5_annual")
    
    site <- pm_stats$station_name[pm_data$ems_id == emsid][1]
    
    ## Subset daily, caaqs, and annual data
    daily_plot <- stnplots[[emsid]]$daily
    annual_plot <- stnplots[[emsid]]$annual
    
    cat("\\subsubsection*{\\underline{", knitr_latex_char(site), " monitoring station}}", sep = "")
    
    if (nrow(caaqs_data_24h)) {
      
      cat("\\textbf{PM$_{2.5}$ 24-hour Air Quality Standard:} ",
          as.character(caaqs_data_24h$caaqs), " \\newline ")
      cat("\\textbf{PM$_{2.5}$ 24-hour Metric:} ", caaqs_data_24h$metric_value,
          " µg/m³ (", caaqs_data_24h$n_years,
          " year average) \\newline ", sep = "")
      plot(set_label_size(daily_plot, 3))
      
    } else {
      cat("\\textbf{PM$_{2.5}$ 24-hour Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newline ")
    cat("\\newline ")
    cat("\\newline ")
    
    if (nrow(caaqs_data_annual)) {
      
      cat("\\textbf{PM$_{2.5}$ Annual Air Quality Standard:} ",
          as.character(caaqs_data_annual$caaqs), " \\newline ")
      cat("\\textbf{PM$_{2.5}$ Annual Metric:} ", caaqs_data_annual$metric_value,
          " µg/m³ (", caaqs_data_annual$n_years,
          " year average) \\newline ", sep = "")
      plot(set_label_size(annual_plot, 3))
      
    } else {
      cat("\\textbf{PM$_{2.5}$ Annual Air Quality Standard:} Insufficient Data")
    }
    
    cat("\\newpage ")
  }
}

```

